The role of the family and the school in the political socialization of attitudes towards gender equality

### 2022-10-14 paper.Rmd

---
title: El rol de la familia y la escuela en la socialización política de las actitudes
  hacia la igualdad de género
author:
- Daniel Miranda^[Pontificia Universidad Católica de Chile, [damiran1@uc.cl](damiran1@uc.cl)]
- Julio Iturra^[Centro de Estudio de Conflicto y Cohesión Social]
- Anaís Herrera^[Instituto de Estudios Avanzados en Educación]
- Kevin Carrasco^[Universidad de Chile]
output:
  pdf_document:
    toc: no
  bookdown::pdf_document2:
    template: null
    keep_tex: no
    number_sections: no
    toc: no
  html_document:
    toc: yes
    code_folding: hide
    toc_float:
      collapsed: yes
      smooth_scroll: no
      number_sections: yes
editor_options:
  chunk_output_type: console
header-includes:
- \usepackage{times}
- \usepackage{caption}
- \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
- \usepackage{graphicx}
- \usepackage{float}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
geometry: margin=0.78in
bibliography: input/bib/Tolerancia.bib
csl: input/bib/apa6.csl
linkcolor: blue
abstract: La socialización política de actitudes hacia la igualdad de género es un
  elemento clave para la construcción de sociedades más democráticas y tolerantes,
  entendiendo la igualdad de derechos entre hombres y mujeres como un principio básico
  de toda democracia y que deben ser fomentados desde la etapa escolar. Utilizando datos del Estudio de Ciudadanía Escolar 2019   se analizan las actitudes de 629 estudiantes de 2° medio y sus apoderados, pertenecientes
  a 60 escuelas de tres regiones de Chile. Se provee evidencia acerca del rol que
  juegan los recursos socioeconómicos de la familia sobre las actitudes hacia la igualdad
  de género de estudiantes y apoderados. Asimismo, se provee evidencia acerca del
  grado de consistencia de la transmisión intergeneracional de actitudes entre estudiantes
  y apoderados. Finalmente, se muestra la influencia sobre las actitudes de los jóvenes
  de dos características de la escuela, el porcentaje de niñas y la percepción de
  apertura a la discusión en el aula.
---

@@

# Introducción
En la actualidad existe una amplio esfuerzo en el mundo por promover desde edades tempranas el reconocimiento activo de los derechos y libertades fundamentales entre hombres y mujeres [@unesco_Declaracion_1995]  como un principio básico de la democracia A pesar de ello persisten amplias desigualdades entre hombres y mujeres, siendo esto particularmente relevante en sociedades latinoamericanas y caribeñas que aún enfrentan grandes desafíos en términos de desigualdades sociales, políticas y económicas que afectan principalmente a grupos tradicionalmente desfavorecidos, como las mujeres [@barcena_matriz_2016]. En ese contexto, comprender la forma en que las creencias o actitudes hacia la igualdad de género son socializadas es un tópico que no ha recibido suficiente atención, particularmente en contextos como el chileno. Así, la presente investigación busca aportar a la comprensión de los factores que influyen en el desarrollo de actitudes hacia la igualdad de género en jóvenes en edad escolar, indagando principalmente en el rol que cumplen dos agentes de socialización: la familia y la escuela.

A pesar de los esfuerzos de promoción de igualdad de género, amplia evidencia muestra que las desigualdades se sostienen en diversos espacios de la vida pública y privada. [Datos sobre la persistencia de la desigualdad en diferentes espacios: representación política, participación laboral en altos cargos, diferencias de ingreso, distribución de tareas del hogar, brechas educativas STEM, persistencia de actitudes sexistas en la población]. Estas brechas existentes en diversos ámbitos presentan ciertas disparidades. En algunos hay mayor grado de avance en acortar las brechas, como se desprende de las iniciativas orientadas a normar igual remuneración ante iguales funciones (citas); mientras que en otros se observan menores avances, como es el caso de violencia de género o distribución de las tareas de cuidado (citas). Por lo tanto, resulta relevante distinguir entre los diferentes ámbitos al indagar la socialización de las actitudes en cuestión.

[Diferencias individuales]
[Rol de la familia: la familia como agente socializador en general (modelamiento - Bandura, Transmisión normativa - Durkheim; consistencia actitudinal como modelamiento, rol de los recursos]
[Rol de la escuela: transmisión normativa, orientación explícita curricular, escuela como contexto]
[Este estudio: combina diferentes aspectos de la familia y la escuela como agentes de socialización de actitudes hacia la igualdad de género en diversos ámbitos (público-privado). Destacar aquello que puede ser considerado novedoso - avance en conocimiento - por ejemplo, evaluar dos agentes relevantes de manera simultánea, combina dos modelos explicativos (recursos y prácticas de socialización), etc.]
[Objetivos]
Dentro del área de investigación sobre el proceso de socialización de actitudes de jóvenes estudiantes se ha indagado extensamente en las características individuales asociadas con actitudes más positivas hacia la igualdad de derechos. En este sentido, se ha destacado que mujeres y quienes tienen un mayor conocimiento cívico poseen actitudes más positivas hacia la igualdad de género [ej. @sampermans_Teachers_2018a; @schulz_Students_2018a] y hacia la igualdad de derechos en general [ej. @torney-purta_How_2008a; @miranda_Political_2018].

Asimismo, al indagar en el rol que juega la escuela y la familia como agentes de socialización, se han encontrado múltiples estudios que abordan las características de la escuela que influyen en la socialización política de las actitudes de los estudiantes [ej. @isac_Native_2012a; @schulz_Students_2018a; @maurissen_Classroom_2020] y, pese a ser más escasa la literatura al respecto, también es posible encontrar una serie de estudios que abordan el rol de la socialización política familiar. En este último ámbito la investigación se ha enfocado en la transmisión intergeneracional de la desigualdad política [@miranda_Political_2018; @villalobos_School_2018a; @isac_Native_2012a], y en la transmisión intergeneracional de actitudes, tanto actitudes tolerantes y/o prejuiciosas [@meeusen_ParentChild_2015; @miklikowska_parent_2016; @miklikowska_Development_2017], como actitudes hacia la igualdad de género [@spierings_Gender_2015; @cunningham_Influence_2001].

Más específicamente, hasta ahora el estudio del rol de la familia y la escuela como agentes de socialización política ha abarcado distintos aspectos. Por un lado, en relación con el rol de la escuela se ha evidenciado la relevancia de la apertura a la discusión en el aula [@schulz_Students_2018a; @trevino_Influence_2017b], el clima escolar [@maurissen_Classroom_2020; @sampermans_Teachers_2018a], la confianza en el valor de la participación de los estudiantes en la escuela (es decir, la confianza de los estudiantes en que su participación en la escuela puede influir en su organización) [@torney-purta_How_2008a; @degroof_Influence_2008a] y las características de los compañeros que comparten en el aula, en términos de la proporción de estudiantes inmigrantes [@isac_Native_2012a; @miklikowska_Driven_2019], y en relación con las actitudes y prejuicios de los compañeros [@miklikowska_Development_2017; @miklikowska_Driven_2019].

Por otro lado, la literatura sobre la transmisión intergeneracional de las actitudes tolerantes se ha aproximado al fenómeno de dos formas distintas. Mientras algunas investigaciones analizan la importancia de las características de la familia a partir del reporte de los estudiantes, otros estudios la analizan a partir de las respuestas de los apoderados. El primer grupo de investigaciones ha generado evidencias sobre la transmisión intergeneracional de la desigualdad política, destacando el efecto de los recursos socioeconómicos de la familia en las actitudes tolerantes de los estudiantes [@miranda_Political_2018; @isac_Native_2012a]. El segundo grupo de estudios ha abordado la transmisión intergeneracional de las actitudes, constatando que las actitudes tolerantes y/o intolerantes de los padres se asocian con las de sus hijos. La mayoría de los artículos relativos a esta asociación refieren a actitudes intolerantes hacia inmigrantes [@meeusen_ParentChild_2015; @miklikowska_parent_2016], homosexuales, musulmanes y prejuicios sexistas [@meeusen_ParentChild_2015], actitudes antiinmigrantes [@miklikowska_Development_2017; @miklikowska_Driven_2019] y actitudes hacia los inmigrantes en términos generales [@miklikowska_parent_2016]. Mientras que, en menor medida,  existen algunas investigaciones en esta área que han constatado esta asociación en las actitudes hacia la igualdad de género en la esfera pública [@spierings_Gender_2015] y privada [@cunningham_Influence_2001].

Adicionalmente, es relevante destacar que se han generado evidencias sobre el efecto moderador de la composición del aula de clases en la relación entre los prejuicios de los padres hacia inmigrantes y los prejuicios de su hijo, mostrando que este efecto es menos pronunciado entre los estudiantes pertenecientes a aulas donde hay una mayor proporción de estudiantes inmigrantes [@miklikowska_Driven_2019]. Sin embargo, esta asociación no ha sido analizada en relación con las actitudes hacia la igualdad de género.

La presente investigación pretende aportar a las líneas de estudio referentes a la influencia de la familia y la escuela en la socialización de actitudes tolerantes y prejuiciosas, analizando la relación entre las actitudes de los apoderados hacia la igualdad de género y las actitudes de sus hijos hacia la igualdad de género. En referencia a investigaciones anteriores, este estudio aportará nuevos antecedentes en dos direcciones. En primer lugar, se pretende aunar los distintos enfoques utilizados en estudios previos sobre las actitudes hacia la igualdad de género, incorporando en la medición de estas actitudes tanto elementos relativos a la igualdad de derechos y oportunidades entre hombres y mujeres, así como indicadores orientados a comprender las actitudes de los jóvenes hacia la distribución de roles de género en las esferas pública y privada. De este modo se busca aportar con antecedentes sobre la transmisión intergeneracional de las actitudes hacia la igualdad de género abordando distintas dimensiones del constructo a la vez. En segundo lugar, este estudio pretende aportar antecedentes sobre el posible rol mitigador de la escuela en la transmisión intergeneracional de actitudes hacia la igualdad de género, cuestión que no ha sido abordada en los estudios revisados, evaluando si algunas características de la escuela permiten moderar la relación entre los recursos socioeconómicos de la familia y las actitudes de los estudiantes, y la relación entre las actitudes de los apoderados y las actitudes de sus hijos.

Los principales objetivos de esta investigación son: a) evaluar en qué medida los recursos de la familia se relacionan con las actitudes hacia la igualdad de género de los jóvenes y de los apoderados; b) evaluar en qué medida las actitudes de los apoderados hacia la igualdad de género se relacionan con las actitudes de sus hijos hacia la igualdad de género; c) evaluar en qué medida características de la escuela a la que asiste el estudiante (más precisamente, la proporción de niñas y la apertura a la discusión en el aula) se relacionan con sus actitudes hacia la igualdad de género; y finalmente, d) evaluar el rol de estas características de la escuela dentro del proceso de socialización como posibles mitigadores de la trasmisión intergeneracional.

# Actitudes hacia la igualdad de género en jóvenes estudiantes chilenos
La igualdad de género se concreta tanto en el espacio privado como en el espacio público y, “en la medida en que se requiere de un reconocimiento social y de la existencia de ciertas condiciones para su concreción en todos los ámbitos de la vida de las mujeres, constituye un pacto social” [@cepal_Planes_2017, p. 12]. Así, se ha avanzado en la concreción de políticas que establezcan condiciones y oportunidades igualitarias, pero se mantienen problemas para asegurar una correcta incorporación de las mujeres al mundo público y su no discriminación, ya que las mujeres acceden a la esfera pública sin dejar sus responsabilidades en el ámbito privado [@cepal_Planes_2017].

Dicho lo anterior, la familia es el primer régimen político en que se transmiten normas relativas a los roles de género que deberían tener hombres y mujeres en la sociedad [@mayer_Gendered_2004]. Estos roles pueden ser definidos como todas las expectativas y creencias en torno a la conducta  de  hombres y mujeres en las esfera pública y privada [@dottisani_Best_2017,, lo cual se traduce en situaciones de desigualdad que favorecen la discriminación, especialmente para las mujeres [azorinabellan_Actitudes_2017]. Por este motivo, la igualdad de género busca  contextos de acuerdo social y condiciones para el ejercicio de los derechos en todos los ámbitos de la vida social, ya que “la falta de garantía de derechos en algún área de la vida de las mujeres incide negativamente en su autonomía como un todo” [@cepal_Planes_2017, p. 12].

El estudio de las actitudes hacia la igualdad de género ha sido abordado desde distintas perspectivas teóricas, por lo que la evidencia empírica sobre esta temática ha operacionalizado este constructo enfocándose en distintos aspectos. Algunos estudios han comprendido las actitudes hacia la igualdad de género como las actitudes hacia la igualdad de derechos y oportunidades entre hombres y mujeres [@miranda_Political_2018; @dottisani_Best_2017; @sampermans_Teachers_2018a], otras investigaciones han abordado la problemática poniendo el énfasis en las actitudes sexistas y en los estereotipos de género [ej. @azorinabellan_Actitudes_2017; @gutierrez_Heroes_2020; @solbes-canales_Socialization_2020], mientras que otros estudios se han enfocado en la distribución de roles de género en las esferas pública y privada [@kaufman_Enduring_2017; @sapiro_Gender_2001; @yu_Decomposing_2013]. Como se enunció anteriormente, en esta investigación se ha intentado aunar los distintos enfoques, incorporando en la medición de las actitudes hacia la igualdad de género tanto elementos relativos a la igualdad de derechos y oportunidades entre hombres y mujeres, como indicadores orientados a comprender las actitudes de los jóvenes en relación con la distribución de roles en las esferas pública y privada. Para ello, se han distinguido tres dimensiones: (1) igualdad de derechos y oportunidades entre hombres y mujeres; (2) distribución igualitaria de roles en la esfera pública; y (3) distribución igualitaria de roles en la esfera privada. Las dos últimas dimensiones incorporan elementos que en otras investigaciones se han abordado como actitudes sexistas y/o estereotipos de género, pero diferenciándolos entre las esferas pública y privada.

Diversos autores han indagado en los factores que inciden en las actitudes hacia la igualdad de género en la población adulta. Se ha evidenciado que  características individuales como el género, la edad, el nivel educacional y el estatus ocupacional son factores que inciden en estas actitudes [ @kyoore_Educational_2019; @charles_Gender_2020a], . Mientras que otro conjunto de investigaciones ha puesto el énfasis en analizar las actitudes hacia la igualdad de género entre los jóvenes, principalmente estudiantes, en relación con sus características individuales y los procesos de socialización política que influyen en sus actitudes hacia distintos grupos sociales tradicionalmente desfavorecidos en general [@miranda_Political_2018] y hacia la igualdad de género en particular [@sampermans_Teachers_2018a; @azorinabellan_Actitudes_2017; @dottisani_Best_2017]. El presente estudio es parte de este segundo conjunto de investigaciones que aborda las actitudes de jóvenes estudiantes, por lo que en este apartado se profundizará en estos antecedentes.

Las principales características individuales de los estudiantes que han sido indagadas en la literatura previa y que se asocian con sus actitudes tolerantes son el nivel de conocimiento cívico y el género. En primer lugar, diversas investigaciones han evidenciado que existe una asociación positiva entre el nivel de conocimiento cívico y las actitudes de los estudiantes hacia la igualdad de derechos para personas inmigrantes [@torney-purta_How_2008a; @isac_Native_2012a; @degroof_Influence_2008a], hacia la igualdad de género y la igualdad de derechos para todos los grupos étnicos y raciales [@schulz_Students_2018a], hacia la igualdad de derechos y oportunidades entre hombres y mujeres [@sampermans_Teachers_2018a] y hacia la igualdad de derechos para personas homosexuales [@schulz_Percepciones_2018a]. En segundo lugar, varias investigaciones han evidenciado que ser mujer se asocia con tener actitudes más positivas hacia la igualdad de derechos para personas inmigrantes [@torney-purta_How_2008a; @isac_Native_2012a; @miranda_Political_2018], con actitudes más positivas hacia minorías étnicas [@degroof_Influence_2008a], hacia la igualdad de derechos para todos los grupos étnicos y raciales [@schulz_Students_2018a; @miranda_Political_2018], hacia la igualdad de derechos para personas homosexuales [@schulz_Percepciones_2018a] y, en específico, hacia la igualdad de derechos y oportunidades entre hombres y mujeres [@miranda_Political_2018; @dottisani_Best_2017] y hacia un reparto equitativo de roles de género en el espacio público [@sapiro_Gender_2001; @kyoore_Educational_2019; @spierings_Gender_2015] y en el espacio privado [@yu_Decomposing_2013]. En consideración de la evidencia disponible, se ha decidido poner atención en la influencia del género en las actitudes de los estudiantes.l. Por lo tanto, se plantea la siguiente hipótesis:

>__*H1*__: Las niñas presentarán actitudes más positivas hacia la igualdad de género que los niños.

>>>>

---
title: "The role of the family and the school in the political socialization of attitudes towards gender equality"
css: "input/css/custom.css" # custom CSS para html
author:
- name: Daniel Miranda ^[Corresponding author - damiran1@uc.cl]
  affiliation: Pontificia Universidad Catolica de Chile, MIDE UC
  number: 1
- name: Julio Iturra
  affiliation: Bremen International Graduate School of Social Sciences
  number: 2
- name: Anaís Herrera
  affiliation: Instituto de Estudios Avanzados en Educación
  number: 3
- name: Kevin Carrasco
  affiliation: Centro de Estudio de Conflicto y Cohesión Social
  number: 4
# Nota: Autores que comparten filiacion, poner el mismo number y declarar filiación una sola vez. 
abstract: The political socialization of attitudes towards gender equality — understanding equality of rights for men and women as a fundamental principle of any democracy — is a crucial element for constructing more democratic and tolerant societies. Using data from the 2019 School Citizenship Panel (Fondecyt, N°1181239), the attitudes of 629 tenth grade students from 60 schools in three regions of Chile and their parents or legal guardians were analyzed. Evidence shows that the socialization of attitudes towards gender equality in its different dimensions depends mainly on the beliefs that parents/guardians manifest. Although socioeconomic conditions have some influence, they do so indirectly through the beliefs and attitudes of parents. Additionally, the school socializes tolerance by promoting greater openness to discussion but does not modify the effects of the family socialization processes described.
output:
  bookdown::pdf_document2:
    template: null
    toc: false
    keep_tex: false
    number_sections: false 
    pandoc_args:
      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion
  bookdown::html_document2:
    number_sections: false
  bookdown::word_document2:
    reference_docx: input/template.docx
    number_sections: false    
always_allow_html: true    
link-citations: yes
linkcolor: blue                         # enlaces y citas en color azul 
editor_options:
  chunk_output_type: console
header-includes:
  - \usepackage {hyperref}
  - \hypersetup {colorlinks = true, linkcolor = blue, urlcolor = blue}  
  - \usepackage{times}           # Times New Roman
  - \usepackage{caption}
  - \captionsetup[figure]{labelfont={bf},labelformat={default},labelsep=period,name={Figure}}
  - \captionsetup[table]{labelfont={bf},labelformat={default},labelsep=period,name={Table}}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
bibliography: ["input/bib/Tolerancia.bib","input/bib/tolerance.bib"]
csl: input/bib/apa6.csl
geometry: "margin=1in" # márgenes de página
fontsize: 12pt
---

@@

```{css, echo=FALSE}
# /* expandir el espacio del documento*/
body .main-container {
      max-width: 1600px !important;
      width: 1600px !important;
    }
#/* expandir width de la TOC*/
div.tocify {
    width: 22% !important;
    max-width: 331px !important;
    max-height: 102% !important;
}
```

# Introduction

There is an extensive effort in the world to promote the active recognition of fundamental rights and freedoms between men and women from an early age [@unesco_Declaracion_1995] as a basic principle of democracy. Despite this, wide inequalities between men and women persist.  These gaps are particularly relevant in Latin American and Caribbean societies that still face great challenges in terms of social, political and economic inequalities that mainly affect traditionally disadvantaged groups, such as women [@barcena_matriz_2016]. Ample evidence shows that inequalities are sustained in various spaces of public and private life. Women are overrepresented in the lowest income quintiles, face a marked wage gap (which, paradoxically, is accentuated depending on the years of study), are affected by an unequal distribution of the use of time and unpaid work in households (responsible for domestic and care work), are the majority of people in the Latin American region without their own income [@cepal_Panorama_2017], have lower participation in formal long-term paid activities, are less likely to have an old-age pension and, those women who access a pension, receive a lower amount than that received by men [@barcena_matriz_2016]. Additionally, it is relevant to distinguish between the disparities in attitudes towards gender equality in the public and private spheres.   For example, different degrees of progress have been made in bridging the respective gaps as can be seen from the initiatives aimed at regulating equal pay for equal functions (job activities) while in other aspects in the private sphere there has been less progress, such as with gender-based violence or the distribution of domestic activities and care tasks.

In this context, understanding how beliefs or attitudes towards gender equality are socialized is a topic that has not received sufficient attention, particularly in contexts such as Chile. Thus, this paper seeks to understand the factors that influence the development of young people’s attitudes towards gender equality, investigating the role played by two socialization agents — the family and the school.

First, in tolerance research, gender is an inescapable dimension. Various studies on the socialization processes of attitudes of young students have investigated individual characteristics and found  gender to be an aspect systematically linked to more positive attitudes towards equal rights. Women  and those with greater civic knowledge tend to demonstrate  more positive attitudes towards gender equality [e.g. @sampermans_Teachers_2018a; @schulz_Students_2018a] and towards equal rights in general [e.g. @torney-purta_How_2008a; @miranda_Political_2018].

Second, the literature on the intergenerational transmission of tolerant attitudes has approached the phenomenon in two ways. While some studies have analyzed the importance of family characteristics from students’ reports, other studies analyze it from the responses of the parents/guardians. The first group of research studies has generated evidence on the intergenerational transmission of political inequality, highlighting the effect of the family’s socio-economic resources on students’ attitudes of tolerance [@miranda_Political_2018; @isac_Native_2012a]. The second group of studies has addressed the intergenerational transmission of attitudes, noting that parents' tolerant and/or intolerant attitudes are associated with those of their children. Most of the articles related to this association refer to intolerant attitudes towards immigrants [@meeusen_ParentChild_2015; @miklikowska_parent_2016], homosexuals, Muslims and sexist prejudices [@meeusen_ParentChild_2015], anti-immigrant attitudes [@miklikowska_Development_2017; @miklikowska_Driven_2019], attitudes towards immigrants in general terms [@miklikowska_parent_2016], and to a lesser extent, have found this association in attitudes towards gender equality in the public [@spierings_Gender_2015] and private sphere [@cunningham_Influence_2001].

Third, multiple studies have addressed the influence of school contextual characteristics  on  the socialization of students’ attitudes [e.g. @isac_Native_2012a; @schulz_Students_2018a; @maurissen_Classroom_2020]. Evidence shows that openness to classroom discussion [@schulz_Students_2018a; @trevino_Influence_2017b], the school climate [@maurissen_Classroom_2020; @sampermans_Teachers_2018a], confidence in the value of student participation in school [@torney-purta_How_2008a; @degroof_Influence_2008a], the characteristics of classmates, in terms of the proportion of immigrant students [@isac_Native_2012a; @miklikowska_Driven_2019], and in relation to peer attitudes and prejudices [@miklikowska_Development_2017; @miklikowska_Driven_2019] all contribute to attitude development. Additionally, evidence has demonstrated the moderating effect of classroom composition on the relationship between parents’ prejudices towards immigrants and their children’s prejudices, showing that this effect is less pronounced among students in classrooms with a higher proportion of immigrant students [@miklikowska_Driven_2019]. However, this role has not been analyzed concerning attitudes towards gender equality.
This research seeks to contribute to the literature regarding the influence of the family and the school context in the socialization of tolerant and prejudicial attitudes, focusing on the relationship between attitudes towards gender equality of parents and their children.

This study will provide new evidence in two directions analyzed in previous research. First, it seeks to integrate two approaches from previous studies, incorporating measures related to equal rights and opportunities between men and women and considering indicators related to the distribution of gender roles within the public and private spheres. In this regard, information about the intergenerational transmission of attitudes towards gender equality can be analyzed, addressing different dimensions of the construct. Secondly, this study aims to provide evidence on the mitigating role of schools in the intergenerational transmission of attitudes towards gender equality, which has not been addressed in previous studies. In particular, we analyze the moderating role of school characteristics over both the relationship between the socio-economic resources of the family and students’ attitudes and the relationship between the attitudes of parents/guardians and the attitudes of their children.

The main objectives of this research are to assess (a) the extent to which the family’s (economic, cultural) resources are related to attitudes towards gender equality of young people and their parents/guardians; (b) the extent to which the attitudes of parents/guardians towards gender equality are related to their children’s attitudes towards gender equality; (c) the extent to which the characteristics of the school attended by the student (more precisely, the proportion of girls in the classroom and the openness to classroom discussion) are related to their attitudes towards gender equality; and finally, (d) the role of these characteristics of the school within the socialization process as possible mitigators of the intergenerational transmission of attitudes.

# Theoretical and Empirical Background

## Attitudes towards gender equality among young people

Gender equality is embodied both in the private and public space "to the extent that it requires social recognition and certain conditions for its realization in all areas of women’s lives" [@cepal_Panorama_2017, p. 12]. In the private sphere, attitudes towards gender roles refer to gender equality within the home. In contrast, in the public sphere, they are related to attitudes toward gender equality in politics, education, work and business [@dottisani_Best_2017].

The approach to the study of attitudes toward gender equality comes from different theoretical perspectives, so the empirical evidence on this issue has operationalized this construct, focusing on different aspects. Some studies have included attitudes towards gender equality such as attitudes towards equal rights and opportunities for men and women [e.g. @miranda_Political_2018; @dottisani_Best_2017; @sampermans_Teachers_2018a]. Other research has addressed the problem by placing emphasis on sexist attitudes and gender stereotypes [e.g. @azorinabellan_Actitudes_2017; @gutierrez_Heroes_2020; @solbes-canales_Socialization_2020], while other studies have focused on the distribution of gender roles in the public and private spheres [e.g. @kaufman_Enduring_2017; @sapiro_Gender_2001; @yu_Decomposing_2013]. First, as noted this research aims to combine the different approaches used in previous studies on attitudes towards gender equality, incorporating in the measurement of these attitudes both elements relating to equal rights and opportunities for men and women, as well as indicators aimed at understanding the attitudes of young people towards the distribution of gender roles in the public and private spheres. To this end, three dimensions have been distinguished: (1) equal rights and opportunities for men and women; (2) equal distribution of roles in the public sphere; and (3) equal distribution of roles in the private sphere.  The latter two dimensions incorporate elements that other research has addressed as sexist attitudes and/or gender stereotypes in either the public or the private spheres.  Here, we sample attitudes about both spheres to analyze how they may differ.

Several authors have investigated the factors that influence attitudes towards gender equality in the adult population. They have generated evidence on the individual factors that influence these attitudes, such as gender, age, educational level and occupational status [e.g. @kyoore_Educational_2019; @charles_Gender_2020a], as well as on the different characteristics of countries and contexts that influence attitudes towards gender equality in general, such as the degree of democratization of the country, its GDP per capita, the percentage of women’s participation in the country’s workforce [@zagrebina_Attitudes_2020] or their percentage of participation in government [@yu_Decomposing_2013]. Other research studies have placed emphasis on analyzing attitudes towards gender equality among young people, mainly students, in relation to their individual characteristics and the processes of political socialization that influence their attitudes towards different social groups traditionally disadvantaged in general [@miranda_Political_2018] and towards gender equality in particular [@sampermans_Teachers_2018a; @azorinabellan_Actitudes_2017; @dottisani_Best_2017]. The present study is part of this second set of research studies.
The main individual characteristic of students that has been investigated in the previous literature and that is associated with their attitudes of tolerance is gender. Several studies show that being female is associated with having more positive attitudes towards equal rights for immigrants [@torney-purta_How_2008a; @isac_Native_2012a; @miranda_Political_2018], more positive attitudes towards ethnic minorities [@degroof_Influence_2008a], towards equal rights for all ethnic and racial groups [@schulz_Students_2018a; @miranda_Political_2018], towards equal rights for homosexual people [@schulz_Percepciones_2018a] and, specifically, towards equal rights and opportunities for men and women [@miranda_Political_2018; @dottisani_Best_2017], towards an equitable distribution of gender roles in the public sphere [@sapiro_Gender_2001; @kyoore_Educational_2019; @spierings_Gender_2015] and in private sphere as well [@yu_Decomposing_2013]. In consideration of the previous evidence, the following hypothesis is raised:

>__*H1*__: Girls will have more positive attitudes towards gender equality than boys.

## The role of the family and the school in the political socialization of attitudes towards gender equality.

The concept of the family as an agent of political socialization has been analyzed from two theoretical models. The first, known as the Resource Model, is concerned with explaining the differences in students’ attitudes based on the differences in the family’s socio-economic resources. Socioeconomic resources indicate that, for example, by possessing a higher level of education, “we are more likely to believe in democratic values and support practices of the same kind” [@lipset_hombre_1997, p. 50]. This is defined by @bobo_Education_1989 as “cognitive sophistication,” accounting for education as a process by which cognitive knowledge and skills are acquired and tolerance towards different groups or opinions thus increases. In this line, @miranda_Political_2018 when studying the participation of young people in political activities points out that the resource model is strongly associated with the social status of the individual, which refers to the individual’s educational level, income and/or occupation, as well as resources such as time and social and financial capital [@waithaka_family_2014]. Thus, a higher educational level would be related to greater tolerance by way of having more time within the educational system to internalize social norms and values [@weber_educational_2020]. Some authors have shown that the socio-economic resources of the family have a direct influence on attitudes towards equal rights of immigrants [@miranda_Political_2018; @villalobos_School_2018a; @isac_Native_2012a] and on changes in anti-immigrant attitudes of students [@miklikowska_Development_2017], on student attitudes towards equal rights and opportunities for men and women [@miranda_Political_2018; @dottisani_Best_2017] and on the general beliefs of young people regarding appropriate behaviors of men and women in the family [@cunningham_Influence_2001]. It has also been shown that, in the adult population, greater socio-economic resources positively affect attitudes towards an equitable distribution of gender roles in public space [@kyoore_Educational_2019] and in private space [@ojeda_Actitudes_2018; @yu_Decomposing_2013]. Following this evidence, hypothesis 2 is suggested.

>__*H2*__: Students from families with greater socio-economic resources will present more positive attitudes towards gender equality.

The second model, known as the Model of attitudinal congruence or intergenerational transmission of attitudes, addresses the family as an agent of political socialization and suggests consistency of attitudes between parents and children towards traditionally disadvantaged groups. As @bandura_Sociallearning_1969 points out, following the theory of social learning, parental attitudes and behaviors and, specifically, sexual roles are an excellent example of active parental training on the appropriate interests and modes of young children’s behavior. From this intergenerational perspective, evidence indicates that a positive association between parental resources and political tolerance could be passed on to subsequent generations [@brady_Political_2015a]. For example, @miklikowska_parent_2016, in a longitudinal study, notes that parents’ attitudes influence children’s attitudes as they convey their beliefs about prejudice and tolerance towards immigrants.  Furthermore, changes in these parents’ attitudes succeed in predicting changes in teenagers’ prejudices and tolerance towards immigrants. For its part, @spierings_Gender_2015, in a comparative study of young people in Turkey and young Turkish residents in Europe, points out that parental attitudes are one of the strongest explanatory variables of young people’s attitudes towards gender equality in the public space and that there are significant differences in belonging to families with parents who support less gender equality in the public space among young Turkish people in Turkey and those residing in other parts of Europe. At the same time, @cunningham_Influence_2001 points out that mothers’ attitudes towards equality in the distribution of gender roles in the family positively influence the general beliefs of children with respect to roles of men and women within the family.  In line with the above, on the one hand @meeusen_ParentChild_2015 argue that the intergenerational similarity of prejudice towards immigrants, Muslims, homosexuals and women is rooted in the ideological attitudes of parents, and that the transfer of prejudicial attitudes and tolerance towards these groups is accentuated according to the frequency of the discussion on political issues and social problems with their children. On the other hand, @miklikowska_Driven_2019 emphasize that parents affect the formation of anti-immigrant attitudes of adolescents through active long-term socialization and through the management of their children's social environment laying the foundation of prejudice.

Although the family’s socio-economic resources have been noted to influence both the attitudes of parents and the attitudes of children, because of the preponderant role played by parents in the socialization of their children, it is expected that their attitudes will play the defining role in the relationship between the socio-economic resources of the family and the attitudes of students. Therefore, based on these previous studies, the following hypotheses on the relevance of family socialization process emerge:

>__*H3*__: Students with parents who have more positive attitudes towards gender equality will present more positive attitudes towards gender equality.

>__*H4*__: Students from families with greater socio-economic resources will present more positive attitudes towards gender equality.

>__*H4a*__: The relationship between family resources and students’ attitudes towards gender equality will be mediated by the attitudes of their parents/guardians towards gender equality.

## The role of the family and the school in the political socialization of attitudes towards gender equality.

Traditionally, education has been assigned the objective of "arousing and developing in the child certain physical, intellectual 
and moral states, which are demanded both by society taken as a whole and by the particular environment to which the child is destined to belong" [@durkheim_Educacion_1999, p.4]. The socializing function of the school "consists of the development within each individual of those skills and attitudes that constitute the essential requirements for future development in life" [@parsons_clase_1976, p. 65]. In this sense, recent research continues to analyze the role of the school as a formative agent for citizens [@debnam_equity_2014].

In terms of the socializing role of the school, several authors have observed that having an environment open to discussion within the classroom promotes students’ attitudes towards the equal rights of immigrants [@isac_Native_2012a; @degroof_Influence_2008a; @torney-purta_How_2008a] and towards the equal rights of immigrants, ethnic groups and women [@schulz_Students_2018a]. However, @maurissen_Classroom_2020 finds that openness to discussion in the classroom loses relevance to promote attitudes towards the equal rights of immigrants when aspects such as responsiveness to students’ demands or the context of respect are not taken into account. In addition, previous research indicates that students’ trust in the deliberative processes in which they participate and confidence that these discussions can influence the organization and functioning of the school are also associated with positive attitudes towards ethnic minorities [@degroof_Influence_2008a] and towards equal rights for immigrants [@torney-purta_How_2008a]. 
The literature also points out that anti-immigrant attitudes of peers have an effect on the anti-immigrant attitudes of students [@miklikowska_Development_2017] and that changes in these anti-immigrant attitudes of peers over time affect the attitudes of students [@miklikowska_Driven_2019]. Likewise, the composition of the classroom also plays a role in the development of student attitudes of tolerance. More precisely, it has been found that the presence of students of different ethnic and/or racial origins, following what is proposed by the contact theory, generates greater support towards tolerance and equal rights for immigrants [@villalobos_School_2018a; @gorodzeisky_Terms_2009]. Additionally, classrooms with a higher proportion of immigrant students moderate the effect of parental prejudice [@miklikowska_Driven_2019] and moderate the association between not being an immigrant and being in favor of equal rights for immigrants [@isac_Native_2012a]. However, there is not enough evidence on the effect that the proportion of girls in the classroom could have on students’ attitudes towards gender equality; this feature was included in the study by @sampermans_Teachers_2018a, but the association was not found to be significant when controlling for the gender of students and the relationships between students and teachers. Considering that previous research that analyzed the effect of classroom composition refers to the same group in which attitudes towards equal rights are being investigated (e.g. the relationship between the proportion of immigrants and attitudes towards equal rights for immigrants), this study follows the same approach regarding the possible effect of the proportion of girls on attitudes towards gender equality.
